---
title: "NYC Traffic Data Analysis"
output: html_notebook
---

This notebook explores NYC accident/collision data and NYS traffic data. 
The overall question is, are accidents explained simply by the amount of 
traffic? If not, are there any insights we can glean to implement safety 
measures for government officials to implement?

```{r}
library(dplyr) 
library(tidyr) 
library(ggplot2) 
library(lubridate)
```

ETL for NYC Motor Vechicle Collisions data
https://dmv.ny.gov/about-dmv/statistical-summaries

```{r ETL}

collisions = read.csv("./Motor_Vehicle_Collisions_-_Crashes.csv")

# Set crash date data type to date
collisions$CRASH.DATE = as.Date(collisions$CRASH.DATE, "%m/%d/%Y")

# Extract hour of day from crash time
collisions$CRASH.HOUR = sub("\\:.*", "", collisions$CRASH.TIME)

# Inspect date range of data
min(collisions$CRASH.DATE) 
max(collisions$CRASH.DATE)

# Since 2012 and 2021 are not a full year, filter between 2013 and 2019
collisions = collisions %>%
  filter(CRASH.DATE >= "2013-01-01" & CRASH.DATE <= "2020-12-31")

```

EDA of collisions data

```{r}
collisions %>%
  ggplot(aes(x=CONTRIBUTING.FACTOR.VEHICLE.1, y=NUMBER.OF.PERSONS.KILLED)) +
    geom_boxplot()

unique(collisions$CONTRIBUTING.FACTOR.VEHICLE.1)

```

Q: What is the leading contributing factor to deaths?

```{r}
collisions %>%
  filter(!CONTRIBUTING.FACTOR.VEHICLE.1 %in% c("", 
                                               "1", 
                                               "80", 
                                              "Unspecified") 
         & NUMBER.OF.PERSONS.KILLED > 0) %>%
  count(CONTRIBUTING.FACTOR.VEHICLE.1) %>%
  arrange(desc(n)) %>%
  top_n(10) %>%
  ggplot(aes(x=CONTRIBUTING.FACTOR.VEHICLE.1, y=n)) +
    geom_bar(stat='identity') + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Driver inattention is leading factor. Does this happen more different times of day? 
collisions %>%
  filter(CONTRIBUTING.FACTOR.VEHICLE.1 == "Driver Inattention/Distraction" 
         & NUMBER.OF.PERSONS.KILLED > 0) %>%
  select(CRASH.HOUR) %>%
  arrange(as.numeric(CRASH.HOUR)) %>%
  ggplot(aes(x=CRASH.HOUR)) +
    geom_bar()
## More during rush hour but we should normalize for traffic or collisions.

```

Q: Have there been more deaths over time?

```{r}

collisions %>%
  filter(NUMBER.OF.PERSONS.KILLED > 0) %>%
  group_by(month = lubridate::floor_date(CRASH.DATE, "month")) %>%
  summarise(NUMBER.OF.PEDESTRIANS.KILLED=sum(NUMBER.OF.PEDESTRIANS.KILLED, na.rm=TRUE),
            NUMBER.OF.CYCLIST.KILLED=sum(NUMBER.OF.CYCLIST.KILLED, na.rm=TRUE),
            NUMBER.OF.MOTORIST.KILLED=sum(NUMBER.OF.MOTORIST.KILLED, na.rm=TRUE)) %>%
  gather(key="type", value="people", -month) %>%
  ggplot(aes(x=month, y=people)) +
    geom_line(aes(color=type, linetype=type))

```
Q: How about injuries?

```{r}
collisions %>%
  filter(NUMBER.OF.PERSONS.INJURED > 0) %>%
  group_by(month = lubridate::floor_date(CRASH.DATE, "month")) %>%
  summarise(NUMBER.OF.PEDESTRIANS.INJURED=sum(NUMBER.OF.PEDESTRIANS.INJURED, na.rm=TRUE),
            NUMBER.OF.CYCLIST.INJURED=sum(NUMBER.OF.CYCLIST.INJURED, na.rm=TRUE),
            NUMBER.OF.MOTORIST.INJURED=sum(NUMBER.OF.MOTORIST.INJURED, na.rm=TRUE)) %>%
  gather(key="type", value="people", -month) %>%
  ggplot(aes(x=month, y=people)) +
    geom_line(aes(color=type, linetype=type))
```

Q: Are there months of the year where more deaths or injuries happen?

```{r}

collisions %>%
  filter(NUMBER.OF.PERSONS.KILLED > 0) %>%
  group_by(month = lubridate::month(CRASH.DATE)) %>%
  summarise(NUMBER.OF.PEDESTRIANS.KILLED=sum(NUMBER.OF.PEDESTRIANS.KILLED, na.rm=TRUE),
            NUMBER.OF.CYCLIST.KILLED=sum(NUMBER.OF.CYCLIST.KILLED, na.rm=TRUE),
            NUMBER.OF.MOTORIST.KILLED=sum(NUMBER.OF.MOTORIST.KILLED, na.rm=TRUE)) %>%
  gather(key="type", value="people", -month) %>%
  ggplot(aes(x=month, y=people)) +
    geom_line(aes(color=type, linetype=type))

collisions %>%
  filter(NUMBER.OF.PERSONS.INJURED > 0) %>%
  group_by(month = lubridate::month(CRASH.DATE)) %>%
  summarise(NUMBER.OF.PEDESTRIANS.INJURED=sum(NUMBER.OF.PEDESTRIANS.INJURED, na.rm=TRUE),
            NUMBER.OF.CYCLIST.INJURED=sum(NUMBER.OF.CYCLIST.INJURED, na.rm=TRUE),
            NUMBER.OF.MOTORIST.INJURED=sum(NUMBER.OF.MOTORIST.INJURED, na.rm=TRUE)) %>%
  gather(key="type", value="people", -month) %>%
  ggplot(aes(x=month, y=people)) +
    geom_line(aes(color=type, linetype=type))

```

ETL NYS Traffic Data 
https://data.ny.gov/Transportation/Annual-Average-Daily-Traffic-AADT-Beginning-1977/6amx-2pbv

```{r}

traffic = read.csv("./Annual_Average_Daily_Traffic__AADT___Beginning_1977.csv")

min(traffic$Year)
max(traffic$Year)

# Filter the traffic data by the same years as the collisions data and the same Boroughs/Counties
traffic = traffic %>%
  filter(Year >= 2013 & County %in% c("Bronx", "Kings", "New York", "Richmond", "Queens"))

# Add Borough column to match counties to boroughs
traffic = traffic %>%
  mutate(Borough = case_when(
    County == "Bronx" ~ "BRONX",
    County == "Kings" ~ "BROOKLYN",
    County == "New York" ~ "MANHATTAN",
    County == "Richmond" ~ "STATEN ISLAND",
    County == "Queens" ~ "QUEENS"
    ))

```

EDA of traffic data

```{r}

traffic %>%
  group_by(County, Year) %>%
  summarise(Traffic=sum(Count, na.rm=TRUE)) %>%
  ggplot(aes(x=Year, y=Traffic, color=County)) +
  geom_line()
# Interesting spike in 2019. Are there more stations? Ride sharing?

traffic %>%
  group_by(Station.ID, Year) %>%
  count(Station.ID) %>%
  filter(n == 2)
# There are a number of bad station ids with n()=2. Not entirely sure what that means, but
# looking at this directionally.

traffic %>%
  group_by(Year) %>%
  summarise(n())
# Same number of stations year over year, so traffic changes not due to station changes.

```
Q: Are there more accidents and/or traffic over time?

``` {r}

collisions %>%
  filter(BOROUGH != "") %>%
  group_by(BOROUGH, year=lubridate::year(CRASH.DATE)) %>%
  summarise(collision=n()) %>%
  ggplot(aes(x=year, y=collision, color=BOROUGH)) +
    geom_line()
    
traffic %>%
  group_by(Year, Borough) %>%
  summarise(traffic=sum(Count, na.rm=TRUE)) %>%
  ggplot(aes(x=Year, y=traffic, color=Borough)) +
    geom_line()

```

Join the datasets by year and borough/county

```{r}
traffic_collisions = collisions %>%
  filter(BOROUGH != "") %>%
  group_by(Year=lubridate::year(CRASH.DATE), BOROUGH) %>%
  summarise(num.ppl.injured=sum(NUMBER.OF.PERSONS.INJURED, na.rm=TRUE), 
            num.ppl.killed=sum(NUMBER.OF.PERSONS.KILLED, na.rm=TRUE),
            num.peds.injured=sum(NUMBER.OF.PEDESTRIANS.INJURED, na.rm=TRUE),
            num.peds.killed=sum(NUMBER.OF.PEDESTRIANS.KILLED, na.rm=TRUE),
            num.peds.killed=sum(NUMBER.OF.PEDESTRIANS.KILLED, na.rm=TRUE),
            num.cyclists.injured=sum(NUMBER.OF.CYCLIST.INJURED, na.rm=TRUE),
            num.cyclists.killed=sum(NUMBER.OF.CYCLIST.KILLED, na.rm=TRUE),
            num.motorists.injured=sum(NUMBER.OF.MOTORIST.INJURED, na.rm=TRUE),
            num.motorists.killed=sum(NUMBER.OF.MOTORIST.KILLED, na.rm=TRUE),
            num.collisions=n()) %>%
  inner_join(traffic %>%
               group_by(Year, Borough) %>%
               summarise(traffic=sum(Count, na.rm=TRUE), roadLength=sum(Length, na.rm=TRUE)), 
             by=c("Year" = "Year", "BOROUGH" = "Borough"))
```

EDA of joined traffic and collision data

```{r}

## Start here and fix this
traffic_collisions %>%
  ggplot(aes(x=traffic/1000, y=num.collisions, color=BOROUGH)) + 
    geom_point()

# It looks like there is one queens point in 2019 that is an anomaly. Going to remove it.
traffic_collisions %>%
  filter(!traffic == 26695902) %>%
  ggplot(aes(x=traffic/1000, y=num.collisions, color=BOROUGH)) + 
    geom_point()

traffic_collisions %>%
  filter(!traffic == 26695902) %>%
  ggplot(aes(x=traffic/1000, y=num.ppl.injured, color=BOROUGH)) + 
    geom_point()

traffic_collisions %>%
  filter(!traffic == 26695902) %>%
  ggplot(aes(x=traffic/1000, y=num.ppl.killed, color=BOROUGH)) + 
    geom_point()

```

Q: Is number of injuries or deaths a function of traffic?

```{r}

model_killed = lm(num.ppl.killed ~ traffic, data = traffic_collisions)
summary(model_killed)

model_injured = lm(num.ppl.injured ~ traffic, data=traffic_collisions)
summary(model_injured)

# Neither are a function of traffic which is good - something else going on.
```

